---
title: "Data Analysis for the Prosody Project"
author: "Lily Zihui Zhu"
date: "12/31/2021"
output: 
    html_document:
      toc: true
      theme: united
---

## Overview

This document contains all R code used for statistical analysis and to generate visualizations for the Prosody Project. There are 4 phonetic features of interest, namely wh-word $\Delta$ F0, wh-word duration, vowel $\Delta$ F0, vowel duration. The analysis includes:

1. Comparing the three-types of wh-questions (EQs/PQs/FQs)

    * A linear mixed effect model with the formula: Phonetic feature ∼ Question type + (1|Participant ID) + (1|Context ID).

    * Visualizations with pitch contour and duration point and line plots for both wh-words and vowel

2. Comparing wh-word positions (mid/end) and wh-in-situ questions (EQs/PQs)

    * A linear mixed effect model with the formula: Phonetic feature ∼ Question type * wh-word position + (1|Participant ID) + (1|Context ID)

    * Visualizations with pitch contour and duration point and line plots for both wh-words and vowels with sentence-medial/final wh-words

3. Compare the pitch contours of PQs and FQs by wh-word position in PQs.

    * t-tests (1) between PQ-mid's and FQs (2) between PQ-end's and FQs
    
    * Visualizations with pitch contour and duration point and line plots for both wh-words with PQs overall, PQ-mid's, and PQ-end's. 

### 0.0 Load all necessary packages

Including (may contain overlaps): 

report, parameters, dplyr, ggplot2, lsr, lme4, afex, multcomp, emmeans, cowplot, ggpubr, lmerTest, egg. 

```{r setup, include=FALSE}
library(report)
library(parameters)
library(dplyr)
library(ggplot2)
library(lsr)
library(lme4)
library(afex)
library(multcomp)
library(emmeans)
library(cowplot)
library(ggpubr)
library(lmerTest)
library(egg)

setwd("C:/Users/zhuzi/Downloads/Prosody_Project/data_analysis")

knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```

### 0.1 Read in source file

```{r}
# load data
data = read.csv("x19_NA_whword_vowel_f0_duration_data.csv", header = TRUE)

# inspect it
# View(data)
```

## Part 1. EQs vs. PQs vs. FQs

### 1.1 Filter out the fillers

Keep only the three question types of interest.
```{r}
Q3 = (data %>% filter(q_type != "filler"))

# inspect it
# View(Q3)
```

### 1.2 Generate summary statistics

Calculate means and sd's for each question type
```{r}
# calculate means by question type
Q3 %>%
  group_by(q_type) %>%
  summarize(mean(wh_delta_pitch), mean(vowel_delta_pitch), mean(wh_duration), mean(vowel_duration))

# calculate standard deviations by question type
Q3 %>%
  group_by(q_type) %>%
  summarize(sd(wh_delta_pitch), sd(vowel_delta_pitch), sd(wh_duration), sd(vowel_duration))
```

### 1.3 Fit data into linear mixed effect models

#### 1.3.1 For wh-word pitch:
```{r}
# fit the model with participant_ID and stimuli as random effects
wh_pitch_qtype_mixed = lmer(wh_delta_pitch ~ q_type + (1|participant_ID) + (1|stimuli_.), data = Q3)

# see if the random effects are necessary 
ranova(wh_pitch_qtype_mixed)

# get slopes coefficients and p-values (not standardized)
# summary(wh_pitch_qtype_mixed)

# generate summary reports based on summary 
# as.report_table(report(wh_pitch_qtype_mixed))

# get main effect of question type
anova(wh_pitch_qtype_mixed)

# standardize and refit the data for standardized beta
model_parameters(wh_pitch_qtype_mixed, standardize = "refit")

# get pairwise comparison results 
emmeans(wh_pitch_qtype_mixed, pairwise~q_type)

# another way to do pairwise comparison after lmer()
# post.hoc.wh_pitch_qtype_mixed <- glht(wh_pitch_qtype_mixed, linfct = mcp(q_type = 'Tukey'))
# summary(post.hoc.wh_pitch_qtype_mixed)

```

#### 1.3.2 For vowel pitch:
```{r}
# fit the model with participant_ID and stimuli as random effects
vowel_pitch_qtype_mixed = lmer(vowel_delta_pitch ~ q_type + (1|participant_ID) + (1|stimuli_.), data = Q3)

# see if the random effects are necessary 
ranova(vowel_pitch_qtype_mixed)

# get slopes coefficients and p-values (not standardized)
# summary(vowel_pitch_qtype_mixed)

# generate summary reports based on summary 
# as.report_table(report(vowel_pitch_qtype_mixed))

# get main effect of question type
anova(vowel_pitch_qtype_mixed)

# standardize and refit the data for standardized beta
model_parameters(vowel_pitch_qtype_mixed, standardize = "refit")

# get pairwise comparison results 
emmeans(vowel_pitch_qtype_mixed, pairwise~q_type)

# another way to do pairwise comparison after lmer()
# post.hoc.vowel_pitch_qtype_mixed <- glht(vowel_pitch_qtype_mixed, linfct = mcp(q_type = 'Tukey'))
# summary(post.hoc.vowel_pitch_qtype_mixed)

```

#### 1.3.3 For wh-word duration:
```{r}
# fit the model with participant_ID and stimuli as random effects
wh_duration_qtype_mixed = lmer(wh_duration ~ q_type + (1|participant_ID) + (1|stimuli_.), data = Q3)

# see if the random effects are necessary 
ranova(wh_duration_qtype_mixed)

# get slopes coefficients and p-values (not standardized)
# summary(wh_duration_qtype_mixed)

# generate summary reports based on summary 
# as.report_table(report(wh_duration_qtype_mixed))

# get main effect of question type
anova(wh_duration_qtype_mixed)

# standardize and refit the data for standardized beta
model_parameters(wh_duration_qtype_mixed, standardize = "refit")

# get pairwise comparison results 
emmeans(wh_duration_qtype_mixed, pairwise~q_type)

# another way to do pairwise comparison after lmer()
# post.hoc.wh_duration_qtype_mixed <- glht(wh_duration_qtype_mixed, linfct = mcp(q_type = 'Tukey'))
# summary(post.hoc.wh_duration_qtype_mixed)

```

#### 1.3.4 For vowel duration:
```{r}
# fit the model with participant_ID and stimuli as random effects
vowel_duration_qtype_mixed = lmer(vowel_duration ~ q_type + (1|participant_ID) + (1|stimuli_.), data = Q3)

# see if the random effects are necessary 
ranova(vowel_duration_qtype_mixed)

# get slopes coefficients and p-values (not standardized)
# summary(vowel_duration_qtype_mixed)

# generate summary reports based on summary 
# as.report_table(report(vowel_duration_qtype_mixed))

# get main effect of question type
anova(vowel_duration_qtype_mixed)

# standardize and refit the data for standardized beta
model_parameters(vowel_duration_qtype_mixed, standardize = "refit")

# get pairwise comparison results 
emmeans(vowel_duration_qtype_mixed, pairwise~q_type)

# another way to do pairwise comparison after lmer()
# post.hoc.vowel_duration_qtype_mixed <- glht(vowel_duration_qtype_mixed, linfct = mcp(q_type = 'Tukey'))
# summary(post.hoc.vowel_duration_qtype_mixed)

```

### 1.4 Make pitch contour and duration plots

#### 1.4.1 For wh-word: 
```{r}

# get dataframe: "time 0 - start pitch", "time duration - end pitch"
wh_pitch_start_end = rbind(Q3 %>% group_by(q_type) %>% summarise(time = 0, pitch = mean(wh_start_pitch)), Q3 %>% group_by(q_type) %>% summarise(time = mean(wh_duration), pitch = mean(wh_end_pitch)))

wh_pitch_start_end

# make the point and line graph
wh_qtype_graph = ggplot(data = wh_pitch_start_end, mapping = aes(x = time, y = pitch, 
      group = q_type, color = q_type, shape = q_type)) +
  geom_point(stat = "summary", fun = mean, position = "dodge")+
  geom_line(stat='summary', fun='mean') +
  ylab("F0(Hz)") + ylim(150,240) +
  xlab("T(s)")+ xlim(0.0,0.4) +
  labs(color = "Type",shape = "Type")
```

#### 1.4.2 For vowel: 

```{r}
# get dataframe: "time 0 - start pitch", "time duration - end pitch"
vowel_pitch_start_end = rbind(Q3 %>% group_by(q_type) %>% summarise(time = 0, pitch = mean(vowel_start_pitch)), Q3 %>% group_by(q_type) %>% summarise(time = mean(vowel_duration), pitch = mean(vowel_end_pitch)))

vowel_pitch_start_end

# make the point and line graph
vowel_qtype_graph = ggplot(data = vowel_pitch_start_end, mapping = aes(x = time, y = pitch,
                            group = q_type, color = q_type, shape = q_type)) +
  geom_point(stat = "summary", fun = mean, position = "dodge")+
  geom_line(stat='summary', fun='mean') +
  ylab("F0(Hz)") + ylim(150,240) +
  xlab("T(s)")+ xlim(0.0,0.4) +
  labs(color = "Type",shape = "Type")
```

#### 1.4.3 Put the two graphs side-by-side: 
```{r results='hide'}
# combine the two graphs on one line with common legend and labels
qtype_graph = ggpubr::ggarrange(wh_qtype_graph, vowel_qtype_graph, common.legend = TRUE,
                        legend = "right", labels = "auto")

# save the graph
# ggsave("qtype_graph_1x2.png", plot = qtype_graph, device = "png", scale = 5)

qtype_graph
```


## Part 2. EQs/PQs vs. mid/end

### 2.1 Filter out FQs
```{r}
# only keep EQs and PQs
EQ_PQ = data %>% filter(q_type %in% c("EQ", "PQ"))

# inspect it
# View (EQ_PQ)
```

### 2.2 Generate summary statistics
```{r}
# get mean by question type
EQ_PQ %>%
  group_by(q_type) %>%
  summarize(mean(wh_delta_pitch), mean(vowel_delta_pitch), mean(wh_duration), mean(vowel_duration))

# get mean by wh-word position
EQ_PQ %>%
  group_by(wh_position) %>%
  summarize(mean(wh_delta_pitch), mean(vowel_delta_pitch), mean(wh_duration), mean(vowel_duration))

# get mean grouped by both question type and wh-word position
EQ_PQ %>%
  group_by(wh_position, q_type) %>%
  summarize(mean(wh_delta_pitch), mean(vowel_delta_pitch), mean(wh_duration), mean(vowel_duration))
```

### 2.3 Fit data into linear mixed effect models

Question type and wh-word position are both entered fixed effects.

#### 2.3.1 For wh-word pitch:

```{r}
# fit the model with participant_ID and stimuli as random effects
wh_pitch_inter_mixed =lmer(wh_delta_pitch ~ q_type*wh_position + (1|participant_ID) + (1|stimuli_.), data = EQ_PQ)

# see if the random effects are necessary
ranova(wh_pitch_inter_mixed)

# get slopes coefficients and p-values (not standardized)
# summary(wh_pitch_inter_mixed)

# generate summary reports based on summary 
# as.report_table(report(wh_pitch_inter_mixed))

# standardize and refit the data for standardized beta
model_parameters(wh_pitch_inter_mixed, standardize = "refit")
```

#### 2.3.2 For vowel pitch:

```{r}
# fit the model with participant_ID and stimuli as random effects
vowel_pitch_inter_mixed =lmer(vowel_delta_pitch ~ q_type*wh_position + (1|participant_ID) + (1|stimuli_.), data = EQ_PQ)

# see if the random effects are necessary
ranova(vowel_pitch_inter_mixed)

# get slopes coefficients and p-values (not standardized)
# summary(vowel_pitch_inter_mixed)

# generate summary reports based on summary 
# as.report_table(report(vowel_pitch_inter_mixed))

# standardize and refit the data for standardized beta
model_parameters(vowel_pitch_inter_mixed, standardize = "refit")
```

#### 2.3.3 For wh-word duration:

```{r}
# fit the model with participant_ID and stimuli as random effects
wh_duration_inter_mixed =lmer(wh_duration ~ q_type*wh_position + (1|participant_ID) + (1|stimuli_.), data = EQ_PQ)

# see if the random effects are necessary
ranova(wh_duration_inter_mixed)

# get slopes coefficients and p-values (not standardized)
# summary(wh_duration_inter_mixed)

# generate summary reports based on summary 
# as.report_table(report(wh_duration_inter_mixed))

# standardize and refit the data for standardized beta
model_parameters(wh_duration_inter_mixed, standardize = "refit")
```

#### 2.3.4 For vowel duration:

```{r}
# fit the model with participant_ID and stimuli as random effects
vowel_duration_inter_mixed =lmer(vowel_duration ~ q_type*wh_position + (1|participant_ID) + (1|stimuli_.), data = EQ_PQ)

# see if the random effects are necessary
ranova(vowel_duration_inter_mixed)

# get slopes coefficients and p-values (not standardized)
# summary(vowel_duration_inter_mixed)

# generate summary reports based on summary 
# as.report_table(report(vowel_duration_inter_mixed))

# standardize and refit the data for standardized beta
model_parameters(vowel_duration_inter_mixed, standardize = "refit")
```

### 2.4 Make pitch contour and duration plots

#### 2.4.1 Group data by wh-word position 
```{r}
# wh-position = 'mid' only 
mid = EQ_PQ %>% filter(wh_position == 'mid')

# wh-position = 'end' only
end = EQ_PQ %>% filter(wh_position == 'end')
```

#### 2.4.2 For wh-words
```{r}
# for mid position, get dataframe: "time 0 - start pitch", "time duration - end pitch"
wh_mid_start_end = rbind(mid %>% group_by(q_type) %>% summarise(time = 0, pitch = mean(wh_start_pitch)), mid %>% group_by(q_type) %>% summarise(time = mean(wh_duration), pitch = mean(wh_end_pitch)))

# plot for wh-word at mid position
wh_mid_inter_graph = 
  ggplot(data = wh_mid_start_end, mapping = aes(x = time, y = pitch, 
  group = q_type, color = q_type, shape = q_type)) +
  geom_point(stat = "summary", fun = mean, position = "dodge")+
  geom_line(stat='summary', fun='mean') +
  ylab("F0(Hz)") + ylim(150,275) +
  xlab("T(s)")+ xlim(0.0,0.42) +
  ggtitle("wh-word sentence-middle") +
  labs(color = "Type",shape = "Type") +
  theme(text = element_text(size = 7), plot.title = element_text(hjust = 0.5))

# for end position, get dataframe: "time 0 - start pitch", "time duration - end pitch"
wh_end_start_end = rbind(end %>% group_by(q_type) %>% summarise(time = 0, pitch = mean(wh_start_pitch)), end %>% group_by(q_type) %>% summarise(time = mean(wh_duration), pitch = mean(wh_end_pitch)))

# plot for wh-word at end position
wh_end_inter_graph = 
  ggplot(data = wh_end_start_end, mapping = aes(x = time, y = pitch,
  group = q_type, color = q_type, shape = q_type)) +
  geom_point(stat = "summary", fun = mean, position = "dodge")+
  geom_line(stat='summary', fun='mean') +
  ylab("F0(Hz)") +ylim(150,275) +
  xlab("T(s)")+ xlim(0.0,0.42)+
  ggtitle("wh-word sentence-final") +
  labs(color = "Type",shape = "Type") +
  theme(text = element_text(size = 7), plot.title = element_text(hjust = 0.5))
```

#### 2.4.3 For vowels
```{r}
# for mid position, get dataframe: "time 0 - start pitch", "time duration - end pitch"
vowel_mid_start_end = rbind(mid %>% group_by(q_type) %>% summarise(time = 0, pitch = mean(vowel_start_pitch)), mid %>% group_by(q_type) %>% summarise(time = mean(vowel_duration), pitch = mean(vowel_end_pitch)))

# plot for vowel at mid position
vowel_mid_inter_graph = 
  ggplot(data = vowel_mid_start_end, mapping = aes(x = time, y = pitch, group = q_type, color = q_type, shape = q_type)) +
  geom_point(stat = "summary", fun = mean, position = "dodge")+
  geom_line(stat='summary', fun='mean') +
  ylab("F0(Hz)") + ylim(150,275) +
  xlab("T(s)")+ xlim(0.0,0.42) +
  ggtitle("vowel sentence-middle") +
  labs(color = "Type",shape = "Type") +
  theme(text = element_text(size = 7), plot.title = element_text(hjust = 0.5))

# for end position, get dataframe: "time 0 - start pitch", "time duration - end pitch"
vowel_end_start_end = rbind(end %>% group_by(q_type) %>% summarise(time = 0, pitch = mean(vowel_start_pitch)), end %>% group_by(q_type) %>% summarise(time = mean(vowel_duration), pitch = mean(vowel_end_pitch)))

# plot for vowel at end position
vowel_end_inter_graph = 
  ggplot(data = vowel_end_start_end, mapping = aes(x = time, y = pitch, group = q_type, color = q_type, shape = q_type)) +
  geom_point(stat = "summary", fun = mean, position = "dodge")+
  geom_line(stat='summary', fun='mean') +
  ylab("F0(Hz)") +ylim(150,275) +
  xlab("T(s)")+ xlim(0.0,0.42)+
  ggtitle("vowel sentence-final") +
  labs(color = "Type",shape = "Type") +
  theme(text = element_text(size = 7), plot.title = element_text(hjust = 0.5))
```

#### 2.4.4 Put the graphs together (2x2)
```{r results='hide'}
# arrange the 4 plots made above
# plots have same scales and right columns have no y-axes
inter_graph = ggarrange(wh_mid_inter_graph +
            theme(legend.position = "none"), 
          wh_end_inter_graph + 
            theme(axis.text.y = element_blank(),
                  axis.ticks.y = element_blank(),
                  axis.title.y = element_blank(),
                  plot.margin = margin(l = 1)),
          vowel_mid_inter_graph +
            theme(legend.position = "none"), 
          vowel_end_inter_graph + 
            theme(axis.text.y = element_blank(),
                  axis.ticks.y = element_blank(),
                  axis.title.y = element_blank(),
                  plot.margin = margin(l = 1),
                  legend.position = "none"), 
          nrow = 2)
```


## Part 3. Compare PQs (by wh-word position) and FQs

First, extract the data containing all PQs and FQs. 
```{r}
PQ_FQ = data %>% filter(q_type %in% c('FQ','PQ'))
```


### 3.1 t-test between PQ-mid's and FQs

Note that both PQ-mid's and FQs have `wh_position == "mid"`.
```{r}
PQmid_FQ = data %>% filter(q_type %in% c("PQ", "FQ"), wh_position == "mid")
t.test(wh_delta_pitch ~ q_type, data = PQmid_FQ, var.equal = TRUE)
t.test(wh_duration ~ q_type, data = PQmid_FQ, var.equal = TRUE)

```


### 3.2 t-test between PQ-end's and FQs

```{r}
PQend_FQ = rbind(data %>% filter(q_type == "FQ"), 
                 data %>% filter(q_type == "PQ", wh_position == "end"))
t.test(wh_delta_pitch ~ q_type, data = PQend_FQ, var.equal = TRUE)
t.test(wh_duration ~ q_type, data = PQend_FQ, var.equal = TRUE)
```


### 3.3 Make the pitch contour plots (wh-words only)

#### 3.3.1 For PQs (overall) and FQs
```{r}
# for PQs and FQs, get dataframe: "time 0 - start pitch", "time duration - end pitch"
PQ_FQ_start_end = rbind(PQ_FQ %>% group_by(q_type) %>% summarise(time = 0, pitch = mean(wh_start_pitch)), PQ_FQ %>% group_by(q_type) %>% summarise(time = mean(wh_duration), pitch = mean(wh_end_pitch)))

# plot for wh-word at mid position
PQ_FQ_inter_graph = 
  ggplot(data = PQ_FQ_start_end, mapping = aes(x = time, y = pitch, 
  group = q_type, color = q_type, shape = q_type)) +
  geom_point(stat = "summary", fun = mean, position = "dodge")+
  geom_line(stat='summary', fun='mean') +
  ylab("F0(Hz)") + ylim(155,210) +
  xlab("T(s)")+ xlim(0.0,0.42) +
  ggtitle("overall") +
  labs(color = "Type",shape = "Type") +
  theme(text = element_text(size = 7), plot.title = element_text(hjust = 0.5))
```

#### 3.3.2 For PQ-mid's and FQs
```{r}
PQmid_FQ_start_end = rbind(PQmid_FQ %>% group_by(q_type) %>% summarise(time = 0, pitch = mean(wh_start_pitch)), PQmid_FQ %>% group_by(q_type) %>% summarise(time = mean(wh_duration), pitch = mean(wh_end_pitch)))

PQmid_FQ_inter_graph = 
  ggplot(data = PQmid_FQ_start_end, mapping = aes(x = time, y = pitch, 
  group = q_type, color = q_type, shape = q_type)) +
  geom_point(stat = "summary", fun = mean, position = "dodge")+
  geom_line(stat='summary', fun='mean') +
  ylab("F0(Hz)") + ylim(155,210) +
  xlab("T(s)")+ xlim(0.0,0.42) +
  ggtitle("sentence-middle PQs") +
  labs(color = "Type",shape = "Type") +
  theme(text = element_text(size = 7), plot.title = element_text(hjust = 0.5))
```

#### 3.3.3 For PQ-mid's and FQs
```{r}
PQend_FQ_start_end = rbind(PQend_FQ %>% group_by(q_type) %>% summarise(time = 0, pitch = mean(wh_start_pitch)), PQend_FQ %>% group_by(q_type) %>% summarise(time = mean(wh_duration), pitch = mean(wh_end_pitch)))

PQend_FQ_inter_graph = 
  ggplot(data = PQend_FQ_start_end, mapping = aes(x = time, y = pitch, 
  group = q_type, color = q_type, shape = q_type)) +
  geom_point(stat = "summary", fun = mean, position = "dodge")+
  geom_line(stat='summary', fun='mean') +
  ylab("F0(Hz)") + ylim(155,210) +
  xlab("T(s)")+ xlim(0.0,0.42) +
  ggtitle("sentence-final PQs") +
  labs(color = "Type",shape = "Type") +
  theme(text = element_text(size = 7), plot.title = element_text(hjust = 0.5))
```

#### 3.3.4 Put the three graphs side-by-side: 
```{r results='hide'}
PQ_FQ_graph = ggarrange(PQ_FQ_inter_graph +
            theme(legend.position = "none"), 
          PQmid_FQ_inter_graph + 
            theme(axis.text.y = element_blank(),
                  axis.ticks.y = element_blank(),
                  axis.title.y = element_blank(),
                  plot.margin = margin(l = 1),
                  legend.position = "none"),
          PQend_FQ_inter_graph + 
            theme(axis.text.y = element_blank(),
                  axis.ticks.y = element_blank(),
                  axis.title.y = element_blank(),
                  plot.margin = margin(l = 1)), 
          nrow = 1)
```
